---
title: "DATA 624 HW2"
author: "Bharani Nittala"
date: "February 12, 2023"
output:
      html_document:
        toc: yes
        toc_float: yes
        theme: yeti
        highlight: kate
        font-family: "Arial"
        code_folding: hide
---

This document contains the homework problems for the Data 624 course. 
Link: https://otexts.com/fpp3/graphics-exercises.html

# Homework 2: Decomposition
## Instructions
Do exercises 3.1, 3.2, 3.3, 3.4, 3.5, 3.7, 3.8 and 3.9 from [Forecasting: Principles and Practice](https://otexts.com/fpp3/) book. Please submit both your Rpubs link as well as attach the .rmd file with your code.

## Packages
```{r message=FALSE, warning=FALSE, class.source = 'fold-show'}
if (!require('fpp3')) (install.packages('fpp3'))
if (!require('latex2exp')) (install.packages('latex2exp'))
if (!require('dplyr')) (install.packages('dplyr'))
if (!require('forecast')) (install.packages('forecast'))
if (!require('seasonal')) (install.packages('seasonal'))
```


## Exercises

### 3.1
Consider the GDP information in `global_economy`. Plot the GDP per capita for each country over time. Which country has the highest GDP per capita? How has this changed over time?

```{r message=FALSE, warning=FALSE}
head(global_economy)
```


```{r message=FALSE, warning=FALSE}
global_economy %>%
  autoplot(GDP/Population, show.legend =  FALSE) +
    labs(title = "GDP per Capita",
       subtitle = "1960-2017",
       x = "Year",
       y = "GDP per capita")
```

```{r message=FALSE, warning=FALSE}
global_economy <- global_economy %>% 
  mutate(GDP_per_capita = GDP/Population) 

global_economy %>%
  filter(GDP_per_capita > 100000) %>%
  autoplot(GDP_per_capita) +
  labs(title= "GDP per capita",
       subtitle = "1960-2017",
       y = "USD")
```

```{r message=FALSE, warning=FALSE}
z <- global_economy %>%
  group_by(Country, GDP, Population) %>%
  summarise(GD_PoP = GDP/Population) %>% 
  arrange(desc(GD_PoP))
head(z)

global_economy %>%
  tsibble(key = Code, index = Year)%>%
  filter(Country=="Monaco") %>%
  autoplot(GDP/Population)
```

Monaco has the highest GDP per capita.


### 3.2
For each of the following series, make a graph of the data. If transforming seems appropriate, do so and describe the effect.

-  United States GDP from `global_economy`.
-  Slaughter of Victorian “Bulls, bullocks and steers” in `aus_livestock`.
-  Victorian Electricity Demand from `vic_elec`.
-  Gas production from `aus_production`.

**United States GDP from** `global_economy`
```{r message=FALSE, warning=FALSE}
global_economy %>% 
  filter(Country == "United States") %>% 
  autoplot() +
  labs(title= "United States GDP",
       subtitle = "1960-2017",
       y = "USD")
```

**Slaughter of Victorian “Bulls, bullocks and steers” in** `aus_livestock`.

```{r message=FALSE, warning=FALSE}
aus_livestock %>%
  filter(Animal == "Bulls, bullocks and steers", State == "Victoria") %>%
  autoplot(Count) +
  labs(title= "Slaughter of Victorian “Bulls, bulls and steers [Monthly]", y = "The Total Count")
```

The trend has come down over time in the last 40 years where there is  downward trend in slaughter of Victorian bulls, bulls, and steers. Interesting to see peculiar trend during 2015 where there was reversal of trend again. 

**Victorian Electricity Demand from** `vic_elec`.

```{r message=FALSE, warning=FALSE}
head(vic_elec)
```

```{r message=FALSE, warning=FALSE}
autoplot(vic_elec)+
    labs(title= "Electricity Demand [30 Minute]", 
         subtitle= "Victoria, Australia",
         y = "MW")
```
when viewing the vic_elec demand plot we witness seasonality. The spikes seen are likely summer or winter [extreme months].

```{r message=FALSE, warning=FALSE}
vic_elec %>%
    index_by(Date) %>%
    summarise(Demand = sum(Demand)) -> daily_demand

daily_demand %>% autoplot(Demand) +
    labs(title= "Daily Electricity Demand", 
         subtitle= "Victoria, Australia",
         y = "MW")
```

There is spike in the first few months of each year and another smaller peak towards the middle of each year. 

**Gas production from** `aus_production`

```{r message=FALSE, warning=FALSE}
head(aus_production)
```


```{r message=FALSE, warning=FALSE}
autoplot(aus_production, Gas) +
  labs(title= "Australian Gas Production")
```

```{r message=FALSE, warning=FALSE}
lambda <- aus_production %>%
  features(Gas, features = guerrero) %>%
  pull(lambda_guerrero)
aus_production %>%
  autoplot(box_cox(Gas, lambda)) +
  labs(y = "",
       title= "Australian Gas Production",
       subtitle = latex2exp::TeX(paste0(
         "Transformed gas production with $\\lambda$ = ",
         round(lambda,2))))+
    theme_replace()+
  geom_line(col = "#69b3a2")
```
Using the Guerrero feature, an optimal value of lambda = 0.12 was obtained. As seen above, the transformation has minimized the seasonal variation across the whole series and has provided an almost consistent throughout. Guerrero's (1993) method to select the lambda which minimises the coefficient of variation for subseries of x.

### 3.3
Why is a Box-Cox transformation unhelpful for the `canadian_gas` data?

```{r message=FALSE, warning=FALSE}
head(canadian_gas)
```


```{r}
canadian_gas %>%
  autoplot(Volume) +
  labs(title = "Canadian Gas Production")
```

```{r}
canadian_gas%>%
  summarise(Volume = sum(Volume)) %>% 
  gg_subseries(Volume)
```
  
It doesn't look like the variation increases/decreases with the level of the series.  


### 3.4
What Box-Cox transformation would you select for your retail data (from Exercise 8 in Section 2.10)?

```{r message=FALSE, warning=FALSE}
set.seed(123)

myseries <- aus_retail %>%
  filter(`Series ID` == sample(aus_retail$`Series ID`,1))

head(myseries)
```

```{r message=FALSE, warning=FALSE}
autoplot(myseries, Turnover)
```

```{r message=FALSE, warning=FALSE}
lambda <- myseries %>%
  features(Turnover, features = guerrero) %>%
  pull(lambda_guerrero)
myseries %>%
  autoplot(box_cox(Turnover, lambda)) +
  labs(y = "",
       title = paste("Transformation with lambda = ", round(lambda,2)))
```

The above shows the transformed retail turn over  with the $\lambda$ parameter of 0.22 chosen using the Guerrero method. The transformation shows a more tamed and consistent amplitude in Turnover throughout in comparison to the original plot of the Turnover.


### 3.5
For the following series, find an appropriate Box-Cox transformation in order to stabilize the variance.  

+ Tobacco from `aus_production`  
+ Economy class passengers between Melbourne and Sydney from `ansett`  
+ Pedestrian counts at Southern Cross Station from `pedestrian`

**Tobacco from** `aus_production`

```{r}
head(aus_production)
```

```{r message=FALSE, warning=FALSE}
autoplot(aus_production, (Tobacco)) +
  labs(title = "Tobacco Production")
```

```{r message=FALSE, warning=FALSE}
aus_tobacco <- aus_production %>% 
  select(Quarter, Tobacco)

lambda <- aus_tobacco %>%
  features(Tobacco, features = guerrero) %>%
  pull(lambda_guerrero)

aus_tobacco %>%
  autoplot(box_cox(Tobacco, lambda)) +
  labs(y = "",
       title = latex2exp::TeX(paste0(
         "Transformed Tobacco Production with $\\lambda$ = ",
         round(lambda,2))))
```

Given our $\lambda$ is near 1 with value of 0.93, there is no substantive transformation resulting from the Box-Cox transformation.

**Economy class passengers between Melbourne and Sydney from** `ansett`

```{r}
eco_mel_syd <- ansett %>% 
  filter(Class == "Economy", Airports == "MEL-SYD")
eco_mel_syd %>% autoplot(Passengers)
```

```{r}
lambda <- eco_mel_syd %>%
  features(Passengers, features = guerrero) %>%
  pull(lambda_guerrero)
eco_mel_syd %>%
  autoplot(box_cox(Passengers, lambda)) +
  labs(y = "",
       title = paste("Box-Cox Transformation with lambda = ", round(lambda,2)))
```


**Pedestrian counts at Southern Cross Station from** `pedestrian`

```{r}
head(pedestrian)
```

View unique `Sensor` categories:
```{r}
unique(pedestrian$Sensor)
```

```{r}
scs_pedestrian <- pedestrian %>%
  filter(Sensor == "Southern Cross Station")
```


```{r}
sct_count <- pedestrian %>%
  filter(Sensor == "Southern Cross Station") %>%
  group_by(Sensor) %>%
  index_by(Week = yearweek(Date_Time)) %>%
  summarise(Count = sum(Count))
sct_count %>% autoplot(Count)

```

```{r}
lambda <- sct_count %>%
  features(Count, features = guerrero) %>%
  pull(lambda_guerrero)
sct_count %>%
  autoplot(box_cox(Count, lambda)) +
  labs(y = "",
       title = paste("Box-Cox Transformation with lambda = ", round(lambda,2)))
```
### 3.7
Consider the last five years of the Gas data from `aus_production`.
```{r message=FALSE, warning=FALSE}
gas <- tail(aus_production, 5*4) %>% select(Gas)
```

a.  Plot the time series. Can you identify seasonal fluctuations and/or a trend-cycle?  


```{r message=FALSE, warning=FALSE}
head(gas)
```

```{r message=FALSE, warning=FALSE}
gas <- tail(aus_production, 5*4) %>% select(Gas)

autoplot(gas, Gas)+
  labs(title = "Quarterly Australian Gas Production",
       subtitle = "Q3 2005 - Q2 2010")
```
  
We can observe there is seasonality as evident by a decrease around Q1 and increase around Q3. There is also an upward trend-cycle in gas production. 

b.  Use `classical_decomposition` with `type=multiplicative` to calculate the trend-cycle and seasonal indices.  

```{r message=FALSE, warning=FALSE}
gas %>%
  model(classical_decomposition(Gas, type = "multiplicative")) %>%
  components() %>%
  autoplot() +
  labs(title = "Classical multiplicative decomposition of total AU Gas Production")
```

c.  Do the results support the graphical interpretation from part a?  
Yes, one can clearly see the upward trend and the seasonality in the decomposition components.

d.  Compute and plot the seasonally adjusted data.  

```{r message=FALSE, warning=FALSE}
class_decomp <- gas %>%
  model(
    classical_decomposition(Gas, type = "multiplicative")
  ) %>%
  components()

class_decomp %>% autoplot() +
  labs(title = "Classical multiplicative decomposition of Australia
                  Gas Production")

as_tsibble(class_decomp) %>%
  autoplot(season_adjust) +
  labs(title = "Seasonally Adjusted Data")
```


e.  Change one observation to be an outlier (e.g., add 300 to one observation), and recompute the seasonally adjusted data. What is the effect of the outlier?  

```{r message=FALSE, warning=FALSE}
gas %>%
  mutate(Gas = if_else(Quarter==yearquarter("2007Q2"), Gas + 300, Gas)) %>%
  model(classical_decomposition(Gas, type = "multiplicative")) %>%
  components() %>%
  as_tsibble() %>%
  autoplot(season_adjust) +
  labs(title = 'Seasonally Adjusted Data with 300 added to "2007 Q2"')
```


Interestingly a new seasonal pattern emerged post adding the outlier. It did have a great impact on the overal trend though.

f.  Does it make any difference if the outlier is near the end rather than in the middle of the time series?

```{r message=FALSE, warning=FALSE}
#change one observation to be an outlier
gas_outlier_2 <- gas
gas_outlier_2$Gas[10] <- gas_outlier_2$Gas[10] + 1000

#recompute the seasonally adjusted data

# STL decomposition
dcmp_2 <- gas_outlier_2 %>%
  model(stl = STL(Gas))

#Compute and plot the seasonally adjusted data
components(dcmp_2) %>%
  as_tsibble() %>%
  autoplot(Gas, colour = "gray") +
  geom_line(aes(y=season_adjust), colour = "#0072B2") +
  labs(y = "Gas Production",
       title = "Seasonally Adjusted Australian Gas Production",
       subtitle = "Purposely skewed with one outlier")
```

Adding the outlier to the end vs. adding it to the middle of the series have small difference from one another. The outlier in both cases eliminates the original increasing trend shown in 3.7.d.

### 3.8
Recall your retail time series data (from Exercise 8 in Section 2.10). Decompose the series using X-11. Does it reveal any outliers, or unusual features that you had not noticed previously?

```{r message=FALSE, warning=FALSE}
x11_dcmp <- myseries %>%
  model(x11 = X_13ARIMA_SEATS(Turnover ~ x11())) %>%
  components()
autoplot(x11_dcmp)+
  labs(title =
    "Decomposition of Australian Department Stores Turnover using X-11.")
```
  
Seasonality was much stronger or more variable during the years of 1982 to around 1990, which is quite the opposite of what was previously observed. There appears to be an outlier in terms of retail data when it is usually towards the end of the year that you see bigger turnover.

### 3.9
Figures [3.19](https://otexts.com/fpp3/decomposition-exercises.html#fig:labour) and [3.20](https://otexts.com/fpp3/decomposition-exercises.html#fig:labour2) show the result of decomposing the number of persons in the civilian labour force in Australia each month from February 1978 to August 1995.  

a.  Write about 3–5 sentences describing the results of the decomposition. Pay particular attention to the scales of the graphs in making your interpretation.  

Based on the plots for the number of persons in the civilian labor force in Australia, it presents an overall increasing trend over time, and the seasonal plot says it contains certain amount of seasonal fluctuations.However, the growth does not appear to be a straight line, as an obvious drop shows at around 1992 ~ 1993; the remainder plot also reflected this drop as an outlier pops downward. We could acknowledge from such a situation that there was a large decrease in labor force during that time, and the causation might be national recession or the revolution of gender distribution needed by the labor market.

b.  Is the recession of 1991/1992 visible in the estimated components?  
we noticed a large drop or dip around 1992, which also reflects an outlier pops downward on the remainder plot, during a recession, we expect a huge lost in the population of labor foce; hence the recessio of 1991/1992 is quite visible in the estimated components.

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>